---
title: "Guide to run {verdepcheck} strategies from a docker container"
author: "NEST CoreDev"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guide to run {verdepcheck} strategies from a docker container}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`verdepcheck` can be used within a docker container to consistently check the version dependencies of an R package.
This vignette demonstrates how to run the [`r-verdepcheck-action`](https://github.com/insightsengineering/r-verdepcheck-action) Github workflow locally with a minimal setup.

1. Download the docker image
1. Run the docker container
1. Open RStudio in the browser
1. Run the code below by changing the `github_id` to the repository you want to check

It assumes that you have `docker` installed on your machine.

_Note_: that you can skip the container setup and run the code below in your local R session.

### Pull and run the docker image

The docker image used in this vignette is available on the Github Container Registry (ghcr.io) and is being used by the Insights Engineering R packages (such as `teal` and `tern`).

```bash
docker pull ghcr.io/insightsengineering/rstudio:latest
docker run --rm -p 8787:8787 -e PASSWORD=test -v ./:/workspace/project ghcr.io/insightsengineering/rstudio
```

### Open the RStudio in the browser

Navigate to [http://localhost:8787](http://localhost:8788) and login with the username `rstudio` and password `test`.

### Run the code below inside the container

```{r, eval=FALSE}
# Dependencies -----------------------------------------------------------------

renv::install(c("checkmate", "withr", "testthat"), prompt = FALSE)
renv::install("insightsengineering/verdepcheck", prompt = FALSE)

# Parameters -------------------------------------------------------------------

github_id <- "insightsengineering/teal.code"
ref <- NULL # --branch <name>, tag or NULL
depth <- "--depth 1" # --depth <depth number> or NULL
build_args <- " " # When empty it should have 1 space
r_cmd_check_args <- " " # When empty it should have 1 space
strategy <- "min_isolated" # one of min_isolated, min_cohort, release or max

# Logic to run action (don't change) -------------------------------------------

repo_path <- withr::local_tempfile(pattern = "repo")

checkmate::assert_string(github_id)
checkmate::assert_string(depth, null.ok = TRUE, pattern = "--depth [0-9]+")
checkmate::assert_string(ref, null.ok = TRUE, pattern = "--branch .+")
checkmate::assert_string(build_args)
checkmate::assert_string(r_cmd_check_args)
checkmate::assert_choice(strategy, c("min_isolated", "min_cohort", "release", "max"))

# Clone repository
system2(
  "git",
  args = list(
    "clone", 
    depth,
    ref,
    sprintf("https://github.com/%s", github_id), 
    repo_path
  ) |> purrr::compact() # remove empty arguments if they are NULL
)

setwd(repo_path)

# Prepare arguments for script call
args <- c()
args[1] <- "" # path to file (should be empty for current directory)
args[2] <- build_args # build_args
args[3] <- r_cmd_check_args # check_args
args[4] <- strategy # strategy

# Mock the function inside the script
testthat::with_mocked_bindings(
  code = source("https://raw.githubusercontent.com/insightsengineering/r-verdepcheck-action/main/script.R"),
  commandArgs = function(...) args, 
  .package = "base"
)
```

